\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-02-15 move_schedules.tex
removed lineno since journal does it anyways

We conclude this section with the main theorem of this effort: an output-sensitive bound on the simulation of persistence dynamically.
  \begin{theorem}\label{prop:sparsity_result}
   Given a pair of filtrations $(K,f), (K,f')$ and a decomposition $R = DV$ of $K$, the size of a minimal sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ of cyclic `move' permutations $s_k = (i_k, j_k)$ satisfying: 
   	\begin{equation}\label{eq:rv_seq_theorem}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	can be determined in $O(m \log \log m)$ time and $O(m)$ space, where $R' = D_d V_d$ denotes a valid decomposition of $(K, f')$. 
   	Moreover, if $i_k < j_k$ for all $k \in [d]$, then computing~\eqref{eq:rv_seq_theorem} requires $O(d\kappa)$ column operations, where:	  
   	$$  \quad \kappa = \sum\limits_{k = 1}^d \, (\lvert \mathbb{I}_{k} \rvert + \lvert \mathbb{J}_{k} \rvert) $$
%	$$  \quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert $$
	where the quantities $\lvert \mathbb{I}_{k} \rvert, \lvert \mathbb{J}_{k} \rvert$ of the intermediate $R_k, V_k$ are given in Proposition~\ref{prop:move_st}. Note that since $\kappa$ depends on the sparsity of the intermediate entries $V_1, V_2, \dots, V_d$ and $R_1, R_2, \dots, R_d$, the bound $O(d \kappa)$ is output-sensitive. 
%	and where $\lvert \mathbb{I}_{k} \rvert$ denotes the number of non-zero entries $V_k[i_k, i_k:j_k]$, $\lvert \mathbb{J}_{k}\rvert$ the number of columns satisfying $\mathrm{low}_{R_k} \in [i_k:j_k]$ and $\mathrm{row}_{R_k}(i_k) \neq 0$, and $d = m - \lvert \mathrm{LCS}(K, K') \rvert$.
\end{theorem}
\begin{proof} \normalsize
	Proposition~\ref{prop:schedule_size} yields the necessary conditions for constructing $\mathcal{S}$  with optimal size $d$ in $O(m \log \log m)$ time and $O(m)$. The definition of $\kappa$ follows directly from Algorithm~\ref{alg:mr}. 
\end{proof}

 %% New section
\subsection{Constructing schedules}\label{sec:schedule_construction}
It is clear from Corollary~\ref{cor:expectation} that one may compute the LCS between two permutations $p, q \in S_m$ via the LIS of a single permutation $p^\ast$, and that computation may be carried out in $O(m \log \log m)$ time. 
It is not immediately clear, however, how to obtain a sorting $p \mapsto q$ from a given $\mathcal{L} = \mathrm{LIS}(p')$ in an efficient way. We outline below a simple procedure which constructs such a sorting $\mathcal{S} = (s_1, \dots, s_d)$ in $O(dm\log m)$ time and $O(m)$ space. 
%Alternatively, given $\mathcal{L}$, the algorithm may be executed online (up to $d$ times) in $O(m \log m)$ time and $O(m)$ space.

@@

Moreover, in contrast to $K_\tau$, the Spearman distance can be computed in $O(m)$ time, has a $\ell_1$ interpretation between ranking, and is often used in the well-known \emph{rank aggregation problem}: whereas obtaining a Kemeni optimal aggregation with respect to $K_\tau$ is NP-hard, the optimal such aggregation with respect to $F$ is obtainable in polynomial time~\cite{dinu2006efficient}. 

@@

To adapt $F(\cdot, \cdot)$ to sortings, we decompose the footrule distance additively via the bound:

@@

%% TODO: elaborate on monotonicity, describe segment tree
In light of the discussion above, we propose a heuristic strategy to minimize~\eqref{eq:sorting_additive} which we observed is both efficient to compute and effective in practice.
The heuristic is inspired by the simplicity of computing the Spearman distance between cyclic permutations at the schedule construction phase. Suppose we begin with an array $\mathcal{A}$ of size $m$ which provides $O(1)$ access and modification, initialized with the (signed) \emph{displacement} of every element in $p$ to its corresponding position in $q$. Since the Spearman distance is simply the sum of the absolute value of these displacements, at any point during during the execution of Algorithm~\ref{alg:sorting} we may obtain $F(\bar{p}, q)$ simply by having access to the sum of every entry in $\mathcal{A}$. There are two degrees of freedom in   Algorithm~\ref{alg:sorting}: the first is in the choice of $\sigma \in \mathcal{D}$, and the second is in choosing $j$. If we fix a heuristic for the latter, then we have a set of possible valid moves $S_\mathcal{D}$ induced by each $\sigma \in \mathcal{D}$ to choose from. Observe that each permutation $s_\sigma \in S_\mathcal{D}$ changes the displacement of every symbol in at most three different ways: 
\[
\mathcal{A}(s_\sigma \circ p) = 
\begin{cases} 
	 \mathcal{A}(\sigma) \pm \lvert i - j \rvert & p^{-1}(\sigma) = i \\
	 \mathcal{A}(\sigma) \pm 1 & i < p^{-1}(\sigma) \leq j \\
	 \mathcal{A}(\sigma) & \text{ otherwise }
\end{cases}
\]
%That is, if $s_\sigma$ represents a move permutation moving $\sigma$ from position $i$ to position $j$, then relative to the previous permutation $p$, the displacement of $\sigma$ in $s_\sigma \circ p$ changes by $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. 
%All other symbols are unchanged. 
Thus, if we replace $\mathcal{A}$ with a data structure  supporting $O(\log m)$ access time to aggregate information and $O(\log m)$ modification time ability on $\lvert i - j \vert$ entries simultaneously, we could greedily choose the next permutation $s$ to minimize:
\begin{equation}\label{eq:greedy_step}
	s_{\text{greedy}} = \argmin_{\text{ valid } s_{\sigma \in S_{\mathcal{D}}}} F(s\circ p, \iota)
\end{equation}  
in $O(d\log m)$ time. 
The former problem reduces to the problem of efficiently calculating \emph{prefix sums}, which is easily solved. 
It is not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$---to handle this, note that one can apply constant-factor updates to multiple values in a \emph{implicit treap} data structure in $O(\log m)$ time via \emph{range updates}. 
 % Can actually only achieve O(d^2 log m) it seems
Since single element modifications, removals, and insertions can all be achieved in $O(\log m)$ expected time with such a data structure, we conclude that the greedy approach solving equation~\eqref{eq:greedy_step} requires just $O(d^2 \log m)$ time. 
%Since single element prefix sum modifications can also be achieved $O(\log m)$ time with these data structures, the time to compute $F(\hat{s}_i, \hat{s}_{i+1})$ is bounded above by $O(\log m)$, 
%thus the greedy solution to equation~\ref{eq:schedule_opt_succ} can be computed in $O(d^2 \log m)$ time.

>>>>

We conclude this section with the main theorem of this effort: an output-sensitive bound on the simulation of persistence dynamically.
  \begin{theorem}\label{prop:sparsity_result}
   Given a pair of filtrations $(K,f), (K,f')$ and a decomposition $R = DV$ of $K$, the size of a minimal sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ of cyclic `move' permutations $s_k = (i_k, j_k)$ satisfying: 
   	\begin{equation}\label{eq:rv_seq_theorem}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	can be determined in $O(m \log \log m)$ time and $O(m)$ space, where $R' = D_d V_d$ denotes a valid decomposition of $(K, f')$. 
   	Moreover, if $i_k < j_k$ for all $k \in [d]$, then computing~\eqref{eq:rv_seq_theorem} requires $O(\kappa)$ column operations, where:	  
   	$$  \quad \kappa = \sum\limits_{k = 1}^d \, (\lvert \mathbb{I}_{k} \rvert + \lvert \mathbb{J}_{k} \rvert) $$
%	$$  \quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert $$
	where the quantities $\lvert \mathbb{I}_{k} \rvert, \lvert \mathbb{J}_{k} \rvert$ of the intermediate $R_k, V_k$ are given in Proposition~\ref{prop:move_st}. Note that since $\kappa$ depends on the sparsity of the intermediate entries $V_1, V_2, \dots, V_d$ and $R_1, R_2, \dots, R_d$, the bound $O(\kappa)$ is output-sensitive. 
%	and where $\lvert \mathbb{I}_{k} \rvert$ denotes the number of non-zero entries $V_k[i_k, i_k:j_k]$, $\lvert \mathbb{J}_{k}\rvert$ the number of columns satisfying $\mathrm{low}_{R_k} \in [i_k:j_k]$ and $\mathrm{row}_{R_k}(i_k) \neq 0$, and $d = m - \lvert \mathrm{LCS}(K, K') \rvert$.
\end{theorem}
\begin{proof} \normalsize
	Proposition~\ref{prop:schedule_size} yields the necessary conditions for constructing $\mathcal{S}$  with optimal size $d$ in $O(m \log \log m)$ time and $O(m)$. The definition of $\kappa$ follows directly from Algorithm~\ref{alg:mr}. 
\end{proof}

 %% New section
\subsection{Constructing schedules}\label{sec:schedule_construction}
It is clear from Corollary~\ref{cor:expectation} that one may compute the LCS between two permutations $p, q \in S_m$ via the LIS of a single permutation $p^\ast$, and that computation may be carried out in $O(m \log \log m)$ time. 
It is not immediately clear, however, how to obtain a sorting $p \mapsto q$ from a given $\mathcal{L} = \mathrm{LIS}(p')$ in an efficient way. We outline below a simple procedure which constructs such a sorting $\mathcal{S} = (s_1, \dots, s_d)$ in $O(dm\log m)$ time and $O(m)$ space, or $O(m \log m)$ time and $O(m)$ space per update in the online setting.

@@

Moreover, in contrast to $K_\tau$, the Spearman distance can be computed in $O(m)$ time and is often used in the well-known \emph{rank aggregation problem}. Indeed, whereas obtaining a Kemeni optimal aggregation with respect to $K_\tau$ is NP-hard, the optimal such aggregation with respect to $F$ is obtainable in polynomial time~\cite{dinu2006efficient}. 

@@

To adapt $F(\cdot, \cdot)$ to sortings, we decompose the Spearman distance additively via the bound:

@@

%% TODO: elaborate on monotonicity, describe segment tree
In light of the discussion above, we propose a heuristic strategy to minimize~\eqref{eq:sorting_additive} which we observed is both efficient to compute and effective in practice.
The heuristic is inspired by the simplicity of computing the Spearman distance between cyclic permutations at the schedule construction phase. Suppose we begin with an array $\mathcal{A}$ of size $m$ which provides $O(1)$ access and modification, initialized with the (signed) \emph{displacement} of every element in $p$ to its corresponding position in $q$. Since the Spearman distance is simply the sum of the absolute value of these displacements, at any point during during the execution of Algorithm~\ref{alg:sorting} we may obtain $F(\cdot, q)$ simply by having access to the sum of every entry in $\mathcal{A}$. At every step, there are only two degrees of freedom in Algorithm~\ref{alg:sorting}: the choice of $\sigma \in \mathcal{D}$ to move, and choice of the target index $j$ to move $\sigma$ to. If we fix a heuristic for the latter, then each $\sigma \in \mathcal{D}$ induces a set of possible valid moves satisfying~\eqref{eq:valid_move}, which we denote as $S_\mathcal{D}(\sigma)$. Now, observe that each permutation $s \in S_\mathcal{D}(\sigma)$ changes the displacement of every symbol in at most three different ways: 
\[
\mathcal{A}(s \circ p) = 
\begin{cases} 
	 \mathcal{A}(\sigma) \pm \lvert i - j \rvert & p^{-1}(\sigma) = i \\
	 \mathcal{A}(\sigma) \pm 1 & i < p^{-1}(\sigma) \leq j \\
	 \mathcal{A}(\sigma) & \text{ otherwise }
\end{cases}
\]
%That is, if $s_\sigma$ represents a move permutation moving $\sigma$ from position $i$ to position $j$, then relative to the previous permutation $p$, the displacement of $\sigma$ in $s_\sigma \circ p$ changes by $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. 
%All other symbols are unchanged. 
Thus, if we replace $\mathcal{A}$ with a data structure  supporting $O(\log m)$ access time to aggregate information and $O(\log m)$ modification time ability on $\lvert i - j \vert$ entries simultaneously, we could greedily choose the next permutation $s$ to minimize:
\begin{equation}\label{eq:greedy_step}
	s_{\text{greedy}} = \argmin_{s \in S_{\mathcal{D}}(\sigma)} F(s\circ p, \iota)
\end{equation}  
in $O(d\log m)$ time. 
The former problem of accessing aggregate information reduces to the problem of efficiently calculating \emph{prefix sums}, which is easily solved. 
It is not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$---however, because we are working with displacements, note the additive modification to each entry in $\lvert i - j \rvert$ is a constant. It is known that one can apply constant-factor updates to multiple values in a \emph{implicit treap} data structure in $O(\log m)$ time via \emph{range updates}~\cite{blelloch1998fast}. 
 % Can actually only achieve O(d^2 log m) it seems
Since single element modifications, removals, and insertions can all be achieved in $O(\log m)$ expected time with such a data structure, we conclude that equation~\eqref{eq:greedy_step} may solved in just $O(d^2 \log m)$ time. 
%Since single element prefix sum modifications can also be achieved $O(\log m)$ time with these data structures, the time to compute $F(\hat{s}_i, \hat{s}_{i+1})$ is bounded above by $O(\log m)$, 
%thus the greedy solution to equation~\ref{eq:schedule_opt_succ} can be computed in $O(d^2 \log m)$ time.
